A Decision Procedure for String Constraints with String-Integer Conversion and Flat Regular Constraints

### 2021-05-07 FMCAD2021/3_flatten.tex
section 3

@@

\subsection{String-Number Conversion Function}

@@

When $x$ of {\parseInt} constraints is restricted to the
generic $\alpha$-flat language ($\alpha$ is fixed),
the (\ref{parse}) and (\ref{parseInt-basic}) still hold 
but $w_i(1\le i\le q)$ is known.
However,
by definition,
the generic $\alpha$-flat language is the finite union of all $\alpha$-flat languages,
so we can enumerate all possible values for $w_i$.
In this way,
the problem can still be reduced to the decidability of PA with exponential function (\paexp), i.e.,
\begin{theorem} \label{thm:string-parInt}
If {\paexp} is decidable, then the satisfiability (validity) of string constraints with {\parseInt} in which all string variables 
are ranged over flat strings is decidable. 
\end{theorem}

>>>>

%!TEX root = paper.tex

In this section, we first define the class of string constraints with string-integer conversion, denoted by $\strint$. Then we define the extension of Presburger arithmetic with exponential functions, denoted by $\paexp$. Finally, we show how to flatten the string constraints in $\strint$ into the arithmetic constraints in $\paexp$.

\subsection{String constraints with string-integer conversion ($\strint$)}

\paragraph{String Terms}
Given a finite alphabet $\Sigma$ and 
a finite set $X$ of string variables over $\Sigma^*$,
we define the set of terms $\textit{Terms}(\Sigma,X)$ 
to be the smallest set satisfying
\begin{itemize}
    \item[1] $\Sigma\cup \{\epsilon\} \cup X \subseteq \textit{Terms}(\Sigma,X)$;
    \item[2] if $t_1,t_2\in \textit{Terms}(\Sigma,X)$, then $t_1 \cdot t_2 \in \textit{Terms}(\Sigma,X)$.
\end{itemize}

We extend $I_X$ to $\textit{Terms}(\Sigma,X)$ by letting $I_X(\epsilon)=\epsilon$, 
for $a\in \Sigma, I_X(a)=a$,
and $I_X(t_1\cdot t_2)= I_X(t_1)\cdot I_X(t_2)$.

\paragraph{String Constraints} \label{par: string constraints}
Given a constraint $\phi$ and an interpretation $I$,
$I\models \phi$ denotes that $I$ satisfies $\phi$,
and $I$ is called a \emph{model} of $\phi$.
We use $\lVert \phi\rVert$ to denote the set of all models of $\phi$.

We define the following three forms of atomic string constraints:
\begin{itemize}
\item An equality constraint $\phi_e$ is of the form 
$t_1 = t_2$, where $t_1,t_2\in \textit{Terms}(\Sigma,X)$.
We define $\lVert \phi_e \rVert = \{I\mid I(t_1)=I(t_2)\}$.
Inequality constraints can be defined analogously.

\item A regular constraint $\phi_r$ is of the form 
$x\in L(\mathcal{A})$,
where $x\in \mathbb{X}$ and $\mathcal{A}$ is a finite state automaton.
We define $\lVert \phi_r \rVert = \{I\mid I(x)\in L(\mathcal{A})\}$.

\item A length constraint $\phi_l$ is a linear constraint over 
$Z \cup \{|x| \mid x\in X\}$, %the values of $|x|$ for all $x\in X$,
where $|\cdot |$ is the length function.
We define $\lVert \phi_l \rVert = \{I \mid I\models \phi_l \}$.
\end{itemize}

In Section~\ref{sec:string-solving},  
we will introduce a new form of atomic string constraints, i.e., \emph{string-number conversion constraints}.
A string constraint is a Boolean combination of atomic string constraints possibly with quantifications over $X\cup Z$. Other notions are same as in the first-order logic.

Giving a string constraint $\Psi$,
the problem of string constraints solving is to decide 
whether $\lVert \Psi \rVert$ is empty,
if not, to compute an interpretation $I$ that satisfies $\Psi$.

\subsection{An Extension of Presburger Arithmetic with Exponential Functions ($\paexp$)}

For a first order theory $T$,
we say theory $T$ admits quantifier elimination (QE) if for any formula in $T$, 
there is a quantifier-free formula equivalent to it.
It is well-known that if a theory admits QE, 
then it is a decidable theory.

The formal definition of PA is given in section \ref{PA}.
Here we add the ordering predicate $\le$ into the signature,
which can be defined by $x \le y\, \Def \, \exists z. x+z=y$.
However, the theory PA so far does not admit QE, 
for example, consider the formula $\exists x.x = y + y$. 
We augment the theory with countable unary divisible predicates
$n|x$, where $n\in \mathbb{N}$, 
$n|x$ is true if and only if $x\ \text{mod}\ n=0$ holds.
This structure of PA that admits QE is denoted by $(\mathbb{N},+)$.

We then introduce a theory {$\paexp$}, denoted by $(\mathbb{N},+,2^x)$, that we work on.
\begin{definition}
    Let $\mathcal{L}=\{0,1,+,\le,n \, |x(n\in \mathbb{N}),2^x,l_2(x)\}$, 
     $(\mathbb{N},+,2^x)$ be a $\mathcal{L}$-theory that has domain $\mathbb{N}$, where 
    \begin{itemize}
        \item  $2^x$ is interpreted to the exponential function of $2$ over $\mathbb{N}$; 
        \item interpretations of $0,1,+,\le,=$ are consistent with PA;
        \item for $n\ge 1$,$n|x$ holds iff $\exists y.x=ny$;
        \item $2^0=1$, for $n \ge 1, 2^n = 2^{n-1}+2^{n-1}$;
        We further assume that if $m,n\in \mathbb{N}, m\le n$, 
        then $2^{m-n} = 1$
        \item $l_2(0)=0$; for $n \ge 1,l_2(n) = y$ iff $2^y \le n < 2^{y+1}$;  $l_2(m-n) = 0$
        if $m,n\in \mathbb{N}$ and $m\le n$. 
    \end{itemize}
\end{definition}

$\lambda_2(x) = 2^{l_2(x)}$ can be defined by $l_2(x)$,
intuitively, $\lambda_2(x)$ means the maximal power of 2 that is not larger than $x$.
Then we have $\lambda_2(x) \le x \le 2\lambda_2(x)-1$,
which will be useful in our proof.

\subsection{Flattening $\strint$ into $\paexp$}

The flattening technique was first introduced 
in \cite{Abdulla 2017}.
Fix an alphabet $\Sigma$ and an abstraction parameter $\alpha$, 
for a given atomic string constraint $\phi$, 
flattening $\phi$ with parameter $\alpha$ 
results in a new string constraint $\phi_\alpha$,
such that 
$R(\lVert \phi_\alpha \rVert) = \lVert \phi \rVert \cap \{I \mid \forall x\in X, I(x)\in \mathbb{G}'(\alpha)\}$,
where $R$ is the renaming function with its domain extended to interpretations in the normal manner.
Intuitively, it restricts $\phi$ to interpret 
string variables over $\mathbb{G}'(\alpha)$.

\cite{Abdulla 2017} discussed the flattening of basic 
string constraints including equality, integer, (regular) grammar and transducer constraints. 
For an atomic string constraint $\phi$,
the flattening $\phi_\alpha$ is still an atomic string constraint 
and is Parikh definable,
so its Parikh image can be expressed by a quantifier free PA formula.
Together with the purity condition,
we obtain an existential quantified PA formula $\rho$.
$\rho$ will be sent to a SMT solver,
if the solver returns an solution $\theta$,
then we can construct an interpretation for $\phi_\alpha$ from $\theta$,
otherwise it means $\phi$ is unsatistiable when 
string variables are interpreted  to $\alpha$-flat languages.

Take a regular constraint $\phi = x\in L(A)$ for example,
the flattening of $\phi$ resutls in a new finite state automaton $A'$ over $\Sigma(\alpha)$,
which encodes running $A$ ``in parallel" 
with the generic $\alpha$-flat automaton.
Let $\rho_1$ be the formula describing the Parikh image of $A'$,
which is a formula over variable sets $\Sigma(\alpha)^\#$.
Let $\rho_2$ be the purity condition \eqref{eq:purity}.
Then we obtain the PA formula $\exists (\Sigma(\alpha))^\#. \rho_1 \wedge \rho_2$.
In order to distinguish between different string variables,
we may replace $a(i)^\# \in \Sigma(\alpha)^\#$ by $(x,a(i))^\#$.

Since the structure of a flat automaton is decided 
by its abstraction parameter $\alpha$, 
a Counter-Example Guided Abstraction 
Refinement (CEGAR) framework is designed, which contains both an under- and an 
over-approximation module, to search the possible values of $\alpha$.
The termination for the overall algorithm is not guaranteed.

@@

@@

When $x$ of {\parseInt} constraints is restricted to the
generic $\alpha$-flat language ($\alpha$ is fixed),
the (\ref{parse}) and (\ref{parseInt-basic}) still hold 
but $w_i(1\le i\le q)$ is known.
However,
by definition,
the generic $\alpha$-flat language is the finite union of all $\alpha$-flat languages,
so we can enumerate all possible values for $w_i$.
In this way,
the problem can still be reduced to the decidability of PA with exponential function ($\paexp$), i.e.,
\begin{theorem} \label{thm:string-parInt}
If {$\paexp$} is decidable, then the satisfiability (validity) of string constraints with {\parseInt} in which all string variables 
are ranged over flat strings is decidable. 
\end{theorem}
